The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-06-12 ms/main.tex
model almost finished

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% MODEL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible. Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change. This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}. The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. All parents die.
	\item Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

We assume that dispersal is cost-free, i.e. all emigrating individuals will settle in a patch, leaving the global population size before and after dispersal unchanged. \linelabel{R1-18}\linelabel{R2-14}\chg{Note however, that our methods are in principle also applicable to a scenario where dispersal is associated with a (potentially type-dependent)\footnote{\francois{remove bracket}} cost.} We split the dispersal step into emigration and immigration. We focus on habitat choice during immigration. Individuals have a bias to immigrate into patches of a certain habitat type. The bias \chg{during} immigration \chg{for} new-habitat patches is set to \chg{zero}, without loss of generality.\footnote{\francois{I did not understand why this is needed}} The bias \chg{for immigration to} an old-habitat patch is denoted by $\pi_{i}$, with the index $i$ indicating the wild-type ($w$) or mutant ($m$) bias. For \chg{$\pi_i<0$}, individuals of type $i$ are more likely to settle in new-habitat patches, for \chg{$\pi_i>0$} the reverse is true. For $\pi_i=0$, individuals do not have a preference and dispersal is \chg{unbiased}.

We consider equal and constant emigration rates for both types and habitats throughout the manuscript. Note again that the method would allow for type- and habitat-dependent emigration rates. We denote the probability \chg{that an individual leaves} its natal patch by $m$. 
Then, the probability \chg{that an individual of type $i$} \chg{emigrates (from any habitat type) and immigrates into }the new habitat is given by 
\begin{equation}\label{eq:dispersal}
\begin{aligned}
m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
\end{aligned}
\end{equation}
where $f_{\text{old}}$ is the frequency of old-habitat patches. \linelabel{R1-19}\chg{
	%Since we are working with a time-discrete model the dispersal step needs to be a probability (and not a rate as is common in continuous-time models). 
	We transformed the bias $\pi_i$ which can take any real value to a positive number, $e^{\pi_i}$. This ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.}

From a biological perspective, a number of dispersal schemes are of particular interest. Precisely, we distinguish between the following five scenarios (see also Fig.~\ref{fig:disp_schemes} for an overview):

\begin{itemize}
	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation. \linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega^\text{old}_w>1-r$ and $\omega^\text{old}_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e. a bias towards old-habitat patches. In practice, individuals are thought to use biotic or abiotic cues to preferentially immigrate into habitats where their fecundity is highest when compared to other habitats. This type of dispersal, i.e. matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.

	The same range of parameters ($\pi_w,\pi_m>0$) is obtained when implementing positive density-dependent dispersal \chg{on immigration. I}ndividuals are more likely to immigrate into patches with higher population densities, so both, the wild type and the mutant, will prefer the old habitat.\footnote{\francois{I would turn this around, otherwise this sounds like a fact. `Indeed, if both the wt and the mutant prefer the old habitat, individuals are more likely to immigrate into patches with...}} Highly populated locations can be an indication for a safe shelter, relevant for prey species, and potentially increase the mating success of individuals. This type of positive density-dependent dispersal on immigration (also called conspecific attraction) is for example found in several amphibians, e.g. the salamander species \textit{Mertensiella luschani} \citep{gautier_2006} and \textit{Ambystoma maculatum} \citep{greene_2016} or the frogs \textit{Oophaga pumilio} \citep{folt_2018}.

	\item \textbf{Old-New:} Under this dispersal scheme, individuals tend to immigrate into habitats where they are fitter than the other type. This translates into the wild type preferring old patches, i.e. $\pi_w>0$, while mutant has a bias towards the new habitat, i.e. $\pi_m<0$. 
	Empirical evidence of this mechanism is scarce. It resembles the recently observed dispersal pattern of the ciliates \textit{Tetrahymena thermophila} with a specialist and generalist type \citep{jacob_2018}. The specialist disperses to its preferred habitat while the generalist prefers to immigrate to a suboptimal habitat where it outcompetes the specialist.

	\item \textbf{New-New:} \chg{This dispersal scheme can be\footnote{\francois{is}} motivated by negative density-dependent dispersal on immigration,} whereby individuals are more likely to move to less populated patches. In these locations, resources might be more abundant, intra-specific competition alleviated and the chance of infection transmission decreased, which may compensate for the potentially reduced habitat quality. The corresponding parameter choice in our model is $\pi_w,\pi_m<0$, i.e. both types have a higher likelihood to immigrate to new-habitat patches. \linelabel{R1-23}\chg{We emphasize that population density in the patches changes over time but the biases are constant. We use the comparison with density-dependent dispersal only to motivate our chosen biases during the relevant phase of rescue, i.e. when the population density in new-habitat patches is low.}
	The mutant being initially rare, it is unlikely that the carrying capacity is reached in new-habitat patches -- as assumed throughout the analysis.\footnote{\francois{Unclear to me. Suggestion: `Because the carrying capacity is not typically reached in new patches in the initial phase of rescue, this bias amounts to negative density-dependent dispersal in that phase.}} 
	Various empirical examples of negative density-dependent dispersal exist. Density-dependent immigration effects as described here, are for example found in the damselfish species \textit{Stegastes adustus} \citep{turgeon_2012} and the migratory birds \textit{Setophaga ruticilla} \citep{wilson_2017}. %In both species, immigrants fill the empty space left by dead or emigrated individuals. Notably, in the fish species, negative density-dependent immigration is further enhanced by residents defending their territory against potential immigrants.

	\item \textbf{New-Old:} For completeness, we also consider the scenario in which both types disperse preferentially into the habitat they are relatively less fit in, i.e. we have $\pi_w<0$ and $\pi_m>0$. This dispersal scheme is \chg{vaguely} related to the concept of an `ecological trap': Individuals tend to immigrate into patches that cannot sustain a population, in its most extreme form resulting in the extinction of the species \citep{battin_2004}. \chg{Note however,} that this does not exactly apply to our framework since here the mutant is able to maintain a population in old-habitat patches.\footnote{\francois{Consider removing the last part about the ecological trap.}}

	\item \textbf{Unbiased dispersal (0-0):} Individuals do not have a bias towards any of the two habitats, $\pi_w=\pi_m=0$. Most theoretical results examining the interplay of dispersal and establishment have \chg{used} this dispersal scheme. Therefore, we use it as a benchmark to which we compare the above defined dispersal patterns.
	%
\end{itemize}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{fig1.pdf}
	\caption{\textbf{Parameter sets for the different dispersal schemes.} \small The colors and markers are the same across all figures. The $x$ and $y$ axes show the dispersal bias of the wild type and mutant, respectively, to immigrate into old-habitat patches ($\pi_w$ and $\pi_m$). The markers are located at the parameter values used for the simulations: $\pi_m=\pi_w=0$ for unbiased dispersal (0-0; {\Large $\bullet$}), $\pi_m=\pi_w=0.5$ for a type-independent bias towards old-habitat patches (Old-Old; {\color{blue} $\blacksquare$}), $\pi_m=-0.5,\pi_w=0.5$ for a wild-type preference towards the old habitat, and a mutant-bias to the new habitat (Old-New; {\color{orange}$\blacktriangle$}), $\pi_m=\pi_w=-0.5$ for a type-independent bias towards the new habitat (New-New; {\color{darkgreen}$\blacktriangledown$}), and $\pi_m=0.5,\pi_w=-0.5$ for the mutant-bias towards the old habitat and a preference of the wild type for the new habitat (New-Old; {\color{purple}$\blacklozenge$}). All the results derived below depend continuously on these parameters, so that varying any of the parameters will not result in a sudden change or discontinuity of the corresponding curves. {\color{orange} Do we need this last sentence?\color{blue}No.}}
	\label{fig:disp_schemes}
\end{figure}

@@

All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges. If not stated otherwise, the default parameter values are used for the stochastic simulations.

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Old-habitat dynamics}
\florence{It is not clear whether population size is at Kold or not}
\linelabel{R1-10}\chg{We assume that the populations in old-habitat patches \linelabel{R1-9-3}are at carrying capacity $K_{\text{old}}$, since both types have a positive growth rate in these patches.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega^\text{old}_w$ and $\omega^\text{old}_m$, respectively. We assume that adaptation to the new habitat comes at a cost, such that the mutant is less fit than the wild type in old-habitat patches ($\omega^\text{old}_m<\omega^\text{old}_w$).
%
\linelabel{R2-9}Mutations occur upon reproduction\footnote{\florence{please check my phrasing}}. The probability that the offspring of a wild-type individual is mutant is given by $\theta$. We neglect back-mutation from mutant to wild type.
\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e. we assume that the wild type and the mutant are competitively equivalent. The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by   
\begin{equation}\label{eq:sold}
     K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m}\approx K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
\end{equation}
where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step. Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$. \linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e. $\omega^\text{old}_m<\omega^\text{old}_w$. Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small. This strongly depends on the dispersal scheme and rate. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

%%%%%%%%%%%%%%%%%%%%%%%%%%
% New habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{New habitat dynamics}
In the new habitat and in the absence of immigration, the wild-type population will go extinct. To account for this, we assume that the number of wild-type \linelabel{R1-15}offspring \chg{per parent} in new-habitat patches is a Poisson distributed value with parameter $(1-r)$, where $r$ measures how strongly the wild type is affected by the environmental deterioration. For $r=1$ the wild-type population becomes locally extinct in new-habitat patches after one generation.

The mutant type, in contrast, has a local growth rate larger than one and will (in expectation) grow in numbers. The number of offspring is again Poisson distributed with per-capita rate $(1+a_{\text{new}})$, the local growth rate of the mutant in new-habitat patches.

\linelabel{R1-17}\linelabel{R2-13}\chg{In the following analysis we will assume that the mutant dynamics in new-habitat patches are density-independent. This is a valid assumption as long as the carrying capacity in these patches, $K_{\text{new}}$, is not reached. Of course, in the simulations the carrying capacity might sometimes be reached, especially for large frequencies of old-habitat patches. This assumption on density-independence also restricts our choice for the carrying capacity in the new habitat. If the carrying capacity in new-habitat patches is much smaller than in old-habitat patches, the new habitat can reach its carrying capacity merely due to immigration of wild-type individuals (even after reproduction where the wild-type population is reduced).\footnote{\francois{Rephrase, as this is not clear. What carrying capacity do you choose and whether your approximation holds or not. Maybe: `Here we examine a carrying capacity in new habitats which is half that in old habitats. If the carrying capacity were much smaller, our analytical approximation would break down.}}}

\subsection*{Simulations}
The algorithm implements the life cycle described above. First, a random number of dispersing individuals \chg{from each patch} is drawn from a binomial distribution with success probability $m$ and sample size equal to the \chg{current patch} population size. The dispersing individuals \chg{are pooled and} distributed according to their type and the dispersal pattern. \linelabel{R2-18}\chg{More precisely, we draw a binomial number of individuals of each type to immigrate into an old-habitat patch with probability $\frac{m_i^{\text{old}}}{m}$ (eq.~\eqref{eq:dispersal}) and then distribute these individuals uniformly at random over the old-habitat patches. The same is done with the remaining dispersing individuals who are distributed uniformly into the new-habitat patches.} \linelabel{R2-19}\chg{In each patch, reproduction is simulated by drawing a Poisson distributed number for each type. The mean of this Poisson number is given by the number of individuals of type $i$ in that patch times the mean number of offspring of a single individual of type $i$ which depends on the patch type (old or new).} 
Mutation of the wild-type \linelabel{R2-20}\chg{offspring} \chg{happens} with \chg{probability} $\theta$, implemented using a binomial distribution. Lastly, if necessary, \linelabel{R1-25}\chg{patches are down-regulated back to carrying capacity by sampling individuals uniformly at random without replacement from the population in a patch until the carrying capacity is reached (hypergeometric sampling).}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% MODEL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model and methods}

\subsection*{Main assumptions and life-cycle}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible. Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change. This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}. The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. 
	All parents die at the end of this step.
	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

We split the dispersal step into emigration and immigration. Emigration is unbiased: all individuals have the same probability $m$ of leaving the patch they were born in. We assumed that dispersal biases affect the immigration step. 
We denote by $\pi_{i}$ the bias for immigration to an old-habitat patch, where the index $i$ refers to the type of the dispersing individual ($w$ for wild type, $m$ for mutant). When $\pi_i<0$, individuals of type $i$ are relatively more likely to settle in new-habitat patches than in old-habitat patches; conversely, their bias is towards old-habitat patches when $\pi_i>0$. The case $\pi_i=0$ corresponds to unbiased dispersal. 
For simplicity, we assume that dispersal is cost-free. While local population sizes may be affected by dispersal, the global size of the metapopulation remains the same before and after dispersal. \linelabel{R1-18}\linelabel{R2-14} Note that our methods can readily be applied to costly dispersal (including to costs that differ among wild-type and mutant individuals), and also to type- and habitat-dependent emigration probabilities.

The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is
%
\begin{equation}\label{eq:dispersal}
\begin{aligned}
m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
\end{aligned}
\end{equation}
where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches. The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}

Qualitatively, there are four possible combinations of dispersal biases. We name them according to the preferences of wild type and then of the mutant (e.g., ``Old-New'', wild-type individuals have a bias toward old-habitat patches, and mutant individuals toward new-habitat patches). We add to these four dispersal schemes the case of unbiased dispersal. Fig.~\ref{fig:disp_schemes} provides an overview of the different schemes, together with the parameter values used in the numerical simulations \footnote{(note that all our results depend continuously on the dispersal parameters, so that varying any of the parameters will not result in a sudden change or discontinuity of the corresponding curves; in other words.).\florence{I am not sure whether this should be said somewhere, and where -- maybe discussion?}}

Each of these dispersal schemes can be related to a biological illustration:

\begin{description}
	\item[Old-Old ($\pi_w>0$, $\pi_m>0$)] Both types of individuals have a bias towards old-habitat patches. \linelabel{R1-22}\linelabel{R2-16}If we assume that mutant individuals have a higher fecundity in old-habitat patches than in new-habitat patches (i.e., $\omega_m^{\text{old}} > \omega_m^{\text{new}}$, which is the case in our numerical examples), then this dispersal scheme correspond to biases towards the habitat where individuals have the highest fecundity. This type of dispersal, which can be described as matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.

	Population densities being high in the old-habitat patches, this dispersal scheme can also be interpreted as positive density-dependent immigration. For prey species, highly populated locations can be an indication for a safe shelter, or of a place with numerous mating opportunities. This type of positive density-dependent immigration (also called conspecific attraction) is for example found in several amphibians, e.g. the salamander species \textit{Mertensiella luschani} \citep{gautier_2006} and \textit{Ambystoma maculatum} \citep{greene_2016} or the frogs \textit{Oophaga pumilio} \citep{folt_2018}.

	\item[Old-New ($\pi_w>0$, $\pi_m<0$)] Wild-type individuals preferentially immigrate into old-habitat patches, while mutants prefer new-habitat patches. This corresponds to immigration to patches where the focal type is fitter than the other type (since $\omega_w^{\text{old}} >\omega_m^{\text{old}}$ and $\omega_m^{\text{new}} >\omega_w^{\text{new}}$). 
	A similar dispersal scheme was recently observed for the ciliates \textit{Tetrahymena thermophila} with a specialist and generalist type \citep{jacob_2018}, and where the specialist disperses to its preferred habitat while the generalist prefers to immigrate to a suboptimal habitat where it outcompetes the specialist.

	\item[New-New ($\pi_w<0$, $\pi_m<0$)] \linelabel{R1-23} Both types of individual preferentially immigrate into new-habitat patches. Population densities being on average lower in the new-habitat patches, and in particular, because the carrying capacity is not typically reached in new-habitat patches during the initial phase of evolutionary rescue, this dispersal scheme can be interpreted as negative density-dependent immigration, whereby individuals are more likely to move to less populated patches. In such locations indeed, resources might be more abundant, intra-specific competition alleviated and the chance of infection transmission decreased, which may compensate for the potentially reduced habitat quality.  
	Density-dependent immigration effects as described here, are for example found in the damselfish species \textit{Stegastes adustus} \citep{turgeon_2012} and the migratory birds \textit{Setophaga ruticilla} \citep{wilson_2017}. %In both species, immigrants fill the empty space left by dead or emigrated individuals. Notably, in the fish species, negative density-dependent immigration is further enhanced by residents defending their territory against potential immigrants.

	\item[New-Old ($\pi_w<0$, $\pi_m>0$)] Wild-type individual preferentially immigrate into new-habitat patches, while mutants prefer old-habitat patches. This dispersal scheme is considered mostly for completeness, because it is biologically quite unlikely, although it can be related to the concept of an `ecological trap', wherein individuals tend to immigrate into patches that cannot sustain a population, in its most extreme form resulting in the extinction of the species \citep{battin_2004}.

	\item[Unbiased dispersal (0-0) ($\pi_w=0$, $\pi_m=0$)] Neither typed has a dispersal bias. Most theoretical results examining the interplay of dispersal and establishment have used this dispersal scheme. We therefore use it as a benchmark to which we compare the biased dispersal schemes.
	%
\end{description}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{fig1.pdf}
	\caption{\textbf{Parameter sets and legends for the different dispersal schemes.} \small The colors and markers are the same across all figures. The horizontal axis is the dispersal bias of the wild type, $\pi_w$ (positive values corresponding to preferential immigration into old-habitat patches), and the vertical axis that of the mutant, $\pi_m$. The markers are located at the parameter values used in the simulations.
		%: $\pi_m=\pi_w=0$ for unbiased dispersal (0-0; {\Large $\bullet$}), $\pi_m=\pi_w=0.5$ for a type-independent bias towards old-habitat patches (Old-Old; {\color{blue} $\blacksquare$}), $\pi_m=-0.5,\pi_w=0.5$ for a wild-type preference towards the old habitat, and a mutant-bias to the new habitat (Old-New; {\color{orange}$\blacktriangle$}), $\pi_m=\pi_w=-0.5$ for a type-independent bias towards the new habitat (New-New; {\color{darkgreen}$\blacktriangledown$}), and $\pi_m=0.5,\pi_w=-0.5$ for the mutant-bias towards the old habitat and a preference of the wild type for the new habitat (New-Old; {\color{purple}$\blacklozenge$}). 
		}
	\label{fig:disp_schemes}
\end{figure}

@@

\todo{to be moved} All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges. If not stated otherwise, the default parameter values are used for the stochastic simulations.

\subsection*{Analysis steps}
We decompose our analysis into several steps of increasing complexity. 
\begin{enumerate}
	\item We first consider an environment that is constant over time, and heterogeneous over space, with a fraction $f_{\text{old}}$ of old-habitat patches and $1-f_{\text{old}}$ of new-habitat patches. The population is initiated with wild-type individuals at carrying capacity in old-habitat patches, and at the migration-selection equilibrium $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details), 
	and with a single mutant individual, either in an old- or in a new-habitat patch. There are no further mutations ($\theta = 0$), and we compute the probability of establishment of a mutant lineage. 
	\item We then consider the same environmental setting, but initialize the population with only wild-type individuals. Mutants can appear by mutation during the simulation ($\theta >0$). We compute the probability of adaptation, i.e. that, during a fixed time interval, a mutant appears by mutation and then establishes. 
	\item Finally, we consider the full scenario where each patch degrades one after the other, as described above. The environment is spatially and temporally variable. The population is initialized with only old-habitat patches, all at carrying capacity, with wild-type individuals only. We compute the probability of rescue, i.e. that a mutant appears by mutation and establishes before the population goes extinct. 
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Additional assumptions for the analytical part}
We make a few additional assumptions in the analytical part of our work; these assumptions are relaxed in the stochastic simulations.

Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents. We also consider that mutants are rare compared to wild-type individuals in old-habitat patches. Temporarily ignoring mutations, and denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of mutants in this old-habitat patch at the beginning of the next generation is 
%
\begin{equation}\label{eq:sold}
K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
%
\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
%
= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
\end{equation}
%
The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches. Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$. \hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal. Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough. In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation. Competition is fiercer as carrying capacity is nearer. }\footnote{\florence{I think this is more Results than Model}}. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of mutants in a new-habitat patch at the beginning of the next generation, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is 
%
\begin{equation}\label{eq:defanew}
\omega_{m}^{\text{new}} \widetilde{N}^{\text{new}}_m = \left(1+a_{\text{new}} \right) \widetilde{N}^{\text{new}}_m,
\end{equation}
which defines $a_{\text{new}}$, the per-capita growth rate of mutants in new-habitat patches. This assumption constrains the differences in carrying capacities among old- and new-habitat patches under which our analysis remains valid. In particular, $K_{\text{new}}$ cannot be too much lower than $K_{\text{old}}$, because in this case, the flow of migrants from old- to new-habitat patches would substantially increase local population sizes in new-habitat patches.

\linelabel{R1-17}\linelabel{R2-13} These assumptions are relaxed in our stochastic simulations.
